# Data Aggregation

In this section, we will focus on aggregating the WiFi data to make it more manageable and suitable for analysis. We will start by explaining the steps in detail to perform data aggregation, then create a function to automate the entire process, and finally apply this function to multiple SQLite3 databases in the `ch3` folder.

If you have your own SQLite3 database file, you can use it for this process. If not, you can download our example dataset to follow along with the tutorial.

::: {.callout-note title="Download Example Data"}
If you don't have your own dataset, download our example SQLite3 database file [here](material/ch3/sample.zip).
:::

## Setup

### Loading Packages

Before we start, let’s ensure that we have all the necessary packages installed and loaded. We will use the `pacman` package to manage our package dependencies.

```{r, warning=FALSE, output=FALSE}
# Install pacman package if not already installed
if (!require(pacman)) install.packages("pacman")

# Load the necessary packages using pacman
pacman::p_load(RSQLite, DBI, data.table, lubridate)
```

## Step-by-Step Data Aggregation

### Step 1: Loading the Data

First, we will load the WiFi data from the SQLite3 database.

```{r}
# Establish a connection to the SQLite3 database
conn <- dbConnect(SQLite(), "material/ch3/sample_1.sqlite3")

# Query the WiFi data from the database
wifi_data <- dbGetQuery(conn, "SELECT sensor_name, timestamp, type, subtype, strength AS rssi, source_address, source_address_randomized FROM packets")

# Convert to data.table
setDT(wifi_data)

# Print the first few rows to verify
head(wifi_data)
```

### Step 2: Filtering by Time

To focus on a specific time period for our analysis, we will define a start and end date using the `ymd_hms` function from the `lubridate` package. We will then filter the data to include only rows within our specified datetime range.

```{r}
# Define start and end datetime for filtering
start_date <- ymd_hms("2024-04-09 19:17:00")
end_date <- ymd_hms("2024-04-09 19:20:00")

# Convert timestamp to datetime format and filter data within the specified datetime range
wifi_data_filtered_time <- wifi_data[
  between(ymd_hms(timestamp), start_date, end_date)
]
```

### Step 3: Filtering by Frame Type

Certain `subtypes` of packets may not be relevant for our purposes. We will remove rows where the `subtype` column contains the word "response".

```{r}
# Remove rows where the subtype contains "response"
wifi_data_filtered_frame <- wifi_data_filtered_time[!grepl("response", subtype)]
```

::: {.callout-note collapse="true"}
## Why We Should Remove "Response" Packets
We remove "response" packets because they are sent from APs to devices, not from devices to APs. In the raw WiFi data collected over a month on a university campus (see the table below), "response" packets account for a large proportion (35.3% for "probe-response"). By filtering them out, we focus on the more relevant packets originating from devices, which provide information about the presence and movement of pedestrians.

| Type       | Subtype               | Raw WiFi data Count | Raw WiFi data Proportion |
|------------|-----------------------|---------------------|--------------------------|
| Management | probe-request         | 714,353             | 2.6%                     |
| Management | probe-response        | 9,532,383           | 35.3%                    |
| Management | authentication        | 352,856             | 1.3%                     |
| Management | deauthentication      | 30,765              | 0.1%                     |
| Management | association-request   | 16,043              | 0.1%                     |
| Management | association-response  | 13,621              | 0.1%                     |
| Management | reassociation-request | 44,284              | 0.2%                     |
| Management | reassociation-response| 33,429              | 0.1%                     |
| Management | disassociation        | 24,525              | 0.1%                     |
| Management | action                | 351,477             | 1.3%                     |
| Data       | data                  | 18,546              | 0.1%                     |
| Data       | null                  | 8,716,923           | 32.3%                    |
| Data       | qos-data              | 4,875,257           | 18.1%                    |
| Data       | qos-null              | 2,253,010           | 8.4%                     |
:::

### Step 4: Filtering by Signal Strength

The signal strength (RSSI) of WiFi packets can provide valuable information about the proximity and location of devices. We will filter our data to include only rows where the `rssi` column falls between -30 and -80 dBm.

```{r}
# Filter data to include only rows with signal strength between -30 and -80 dBm
wifi_data_filtered_strength <- wifi_data_filtered_frame[between(rssi, -80, -30)]
```

::: {.callout-note collapse="true"}
## Why We Should Filter Signal Strength
We filter the signal strength to remove extreme values. Very strong signals (above -30 dBm) may come from devices very close to the sensor, like a tablet right next to it, which may not represent typical pedestrian behavior. Very weak signals (below -80 dBm) may be less reliable, coming from far away or obstructed devices. Focusing on the -30 to -80 dBm range helps capture data more likely to represent pedestrians moving within a reasonable distance from the sensors.
:::


### Step 5: Aggregating Data by Time Intervals

We will aggregate the data by `sensor_name` and `source_address` into specified time intervals. This involves rounding the `timestamp` to the nearest interval and then summarizing the data.

```{r}
# Define aggregation function
aggregate_intervals <- function(data, interval) {
  data[, timestamp := floor_date(ymd_hms(timestamp), unit = interval)]
  aggregated_data <- data[, .(
    median_rssi = median(rssi),
    count = .N
  ), by = .(sensor_name, source_address, source_address_randomized, timestamp)]
  return(aggregated_data)
}

# Aggregate the data by 1-second intervals
aggregated_data <- aggregate_intervals(wifi_data_filtered_strength, "second")

# Print the first few rows of the aggregated data
head(aggregated_data)
```

### Step 6: Saving Aggregated Data

Finally, we will save the aggregated data to a CSV file for future use.

```{r}
# Save the aggregated data to a CSV file
fwrite(aggregated_data, "material/ch3/aggregated_sample_1.csv")
```

### Step 7: Closing the Database Connection

After loading and processing the data, it is important to close the database connection to free up resources:

```{r}
dbDisconnect(conn)
```

## Tracking Data Changes

We will create a table that shows how the number of packets and unique `source_address` changes through each step of the data aggregation process.

```{r}
# Initialize summary table
summary_table <- data.table(
  Step = character(),
  Packet_Count = integer(),
  Unique_Source_Addresses = integer()
)

# Initial counts
summary_table <- rbind(summary_table, data.table(
  Step = "Initial",
  Packet_Count = nrow(wifi_data),
  Unique_Source_Addresses = length(unique(wifi_data$source_address))
))

# After filtering by datetime range
summary_table <- rbind(summary_table, data.table(
  Step = "After Datetime Filter",
  Packet_Count = nrow(wifi_data_filtered_time),
  Unique_Source_Addresses = length(unique(wifi_data_filtered_time$source_address))
))

# After filtering by frame type
summary_table <- rbind(summary_table, data.table(
  Step = "After Frame Type Filter",
  Packet_Count = nrow(wifi_data_filtered_frame),
  Unique_Source_Addresses = length(unique(wifi_data_filtered_frame$source_address))
))

# After filtering by signal strength
summary_table <- rbind(summary_table, data.table(
  Step = "After Strength Filter",
  Packet_Count = nrow(wifi_data_filtered_strength),
  Unique_Source_Addresses = length(unique(wifi_data_filtered_strength$source_address))
))

# After aggregation
summary_table <- rbind(summary_table, data.table(
  Step = "After Aggregation",
  Packet_Count = nrow(aggregated_data),
  Unique_Source_Addresses = length(unique(aggregated_data$source_address))
))

# Print the summary table
print(summary_table)
```

Here is the interpretation:

1. **Initial**: The raw data contains 11,490 packets from 323 unique devices.
2. **After Datetime Filter**: Filtering by the specified datetime range reduces the data to 5,274 packets from 163 unique devices.
3. **After Frame Type Filter**: Removing "response" frame types further reduces the data to 3,380 packets from 123 unique devices.
4. **After Strength Filter**: Filtering by signal strength between -30 and -80 dBm results in 2,904 packets from 112 unique devices.
5. **After Aggregation**: Aggregating the data into 1-second intervals leaves us with 522 aggregated data points from the same 112 unique devices.

Each step of the filtering process significantly reduces the amount of data, refining it to include only the most relevant packets and maintaining the same number of unique devices after the final aggregation step.

## Automating the Process with a Function

To streamline the process, we will create a function that performs the entire data aggregation process. This function will:
1. Load the data from the SQLite3 database.
2. Perform basic filtering by datetime, frame type, and signal strength.
3. Aggregate the data by `sensor_name` and `source_address` into specified time intervals.
4. Save the aggregated data to a CSV file.

```{r}
aggregate_data <- function(db_path, start_date, end_date, interval = "second", output_suffix = "_1second.csv") {
  # Establish a connection to the SQLite3 database
  conn <- dbConnect(SQLite(), db_path)
  
  # Query the WiFi data from the database
  wifi_data <- dbGetQuery(conn, "SELECT sensor_name, timestamp, type, subtype, strength AS rssi, source_address, source_address_randomized FROM packets")
  setDT(wifi_data)
  
  # Filter by datetime range
  wifi_data <- wifi_data[between(ymd_hms(timestamp), start_date, end_date)]
  
  # Filter by frame type
  wifi_data <- wifi_data[!grepl("response", subtype)]
  
  # Filter by signal strength
  wifi_data <- wifi_data[between(rssi, -80, -30)]
  
  # Aggregate the data by specified intervals
  wifi_data[, timestamp := floor_date(ymd_hms(timestamp), unit = interval)]
  aggregated_data <- wifi_data[, .(median_rssi = median(rssi), count = .N), by = .(sensor_name, source_address, source_address_randomized, timestamp)]
  
  # Define the output path for the aggregated data
  output_path <- sub("\\.sqlite3$", output_suffix, db_path)
  
  # Save the aggregated data to a CSV file
  fwrite(aggregated_data, output_path)
  
  # Close the database connection
  dbDisconnect(conn)

}

# Example usage
start_date <- ymd_hms("2024-04-09 19:17:00")
end_date <- ymd_hms("2024-04-09 19:20:00")
aggregate_data("material/ch3/sample_1.sqlite3", start_date, end_date, interval = "second")

```

### Applying the Function to Multiple Databases

We will apply the `aggregate_data` function to all SQLite3 databases in the `ch3` folder that contain the word "sample" in their filenames. The resulting CSV files will be saved with "_1second" appended to their original names. Using `lapply` is effective for applying a function over a list, but we can also use `purrr::map` for better readability and functionality.

```{r}
pacman::p_load(purrr)

# List all SQLite3 files in the ch3 folder that contain "sample_"
db_files <- list.files("material/ch3", pattern = "sample_.*\\.sqlite3$", full.names = TRUE)
print(db_files)

# Define the start and end dates for filtering
start_date <- ymd_hms("2024-04-09 19:17:00")
end_date <- ymd_hms("2024-04-09 19:20:00")

# Apply the aggregation function to each database file
map(db_files, ~aggregate_data(.x, start_date, end_date, interval = "second"))
```